Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-02 paper/ms.tex
update ms

\author{coauthors, to include but not limited to
Hillenbrand, Howard, Isaacson, Palumbo}

% <250 words
\begin{abstract}
  %Recent analyses of FGK stars in open clusters have revealed how
  %stellar rotation rates and lithium content evolve over the first few
  %billion years.
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate, and its lithium
  content, can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium
  is measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid; these rotation-based ages are
  roughly an order of magnitude more precise than isochrone ages in
  this sample of FGK stars.
  %
  The independent lithium ages agree with the rotation-based ages for
  $\approx$90\% of cases, suggesting that most of the rotation-based
  ages are not only precise, they are also accurate.
  %However, the timescales of lithium depletion are such that
  %rotation-based ages are more precise for most of our stellar sample.
  %
  Our resulting catalog includes \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  More broadly, we find that the star formation rate in the Kepler
  field has dropped by a factor of \ratiosfr$\pm$\uncratiosfr over the
  past 3\,Gyr, for both planet hosts as well as for the overall
  stellar sample.
  %
  This ``demographic cliff'' has implications for the star formation
  history of our Galaxy, and it helps clarify the relative balance of
  young and old planets that have been detected in transit surveys of nearby
  stars in the Galaxy.
  %
\end{abstract}

@@

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
billions of years old, similar to the Earth.  This fact leaves many
gaps in our knowledge of what exact dynamical and physical processes
produce the observed exoplanet archetypes.    The reason is that
processes such as thermal cooling \citep{2007ApJ...659.1661F},
atmospheric loss \citep{2019AREPS..47...67O}, giant impacts
\citep{2014prpl.conf..595R}, and dynamical instabilities
\citep{2017MNRAS.470.1750I} are expected to be most efficient over
$\approx$10\,Myr to $\approx$1\,Gyr timescales; when we observe most
known exoplanets, these processes are done, and not much is happening.

@@

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 0.5-4\,Gyr, and can also provide useful age limits at earlier
times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
clock is quite old
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  
In the context of Kepler,
empirical models have been previously applied to derive rotation-based
ages for subsets of the Kepler data
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.
There have been major improvements since these pioneering studies in:
\begin{enumerate}[label={\it \roman*)}]
  \item The extent of the Kepler Object of Interest catalog, as well
    as our understanding of which of these objects are false
    positives;
  %
  \item Our empirical understanding of how rotation periods of stars
    in open clusters evolve
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item Our ability to model not only the mean evolution but also the
    intrinsic dispersion in rotation periods as a function of time
    \citep{Bouma_2023}.  
    For Sun-like stars, the new model is broadly consistent with older
    calibrations \citep[e.g.][]{Mamajek_2008}.  However, for late G
    dwarfs and K dwarfs, the age-calibration differs by factors of a
    few;
  %
  \item Our ability to identify binary stars, which can potentially
    problematic for estimates of rotation-based ages.
\end{enumerate}
Today, rotation-based ages for FGK stars below $\approx$4\,Gyr are
empirically tied to the cluster age scale and rotation sequence, and
are limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.

@@

\begin{figure*}[!t]
  \begin{center}
    \subfloat{
      \includegraphics[width=0.49\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
    }
    \subfloat{
    \includegraphics[width=0.49\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
    }
  \end{center}
  \vspace{-0.5cm}
  \caption{
    {\bf The stars.}  Our analysis focuses on stars observed by Kepler
    with previously reported rotation periods (dark gray points).
    About half of these rotators are suitable for gyrochronology
    (black points), based on factors including their temperatures,
    surface gravities, non-binarity, and proximity to the main
    sequence (see Section~\ref{subsec:flags}).   Some are
    ``confirmed'' or ``candidate'' Kepler Objects of Interest that
    meet an additional set of planetary quality criteria (blue points;
    Section~\ref{subsec:plflags}).  Surface gravities and effective
    temperatures are primarily from \citet{Berger_2020a_catalog}.
  }
  \label{fig:stellarprops}
\end{figure*}

\begin{figure*}[!t]
  \begin{center}
    \leavevmode
    \subfloat{
      \includegraphics[width=0.49\textwidth]{gyromodeldispersion.pdf}
    }
    \subfloat{
      \includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf}
    }
  \end{center}
  \vspace{-0.6cm}
  \caption{
    {\bf The models.}
    Points represent $10^4$ draws from model probability distributions
    that have been fitted to observed rotation periods
    \citep{Bouma_2023} and lithium equivalent widths
    \citep[EWs;][]{Jeffries_2023} of stars in open clusters.  Lines
    are the ``mean models'' at various ages; the asymmetric intrinsic
    dispersion in rotation and lithium about these mean models, which
    is what the models fit, sets the theoretical precision limit for
    both age-dating methods.    Additional sources of uncertainty,
    including measurement uncertainty, impose further limits on
    achievable age precision.  These models are calibrated using open
    clusters younger than 4\,Gyr.  The displayed points assume a
    uniform distribution in temperature for visual clarity.  The sizes
    of the points are the same in each panel, so that low apparent
    density signifies greater dispersion around the mean.
    \label{fig:models}
  }
\end{figure*}

\begin{figure*}[!t]
  \begin{center}
    \subfloat{
      \includegraphics[width=0.48\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
    }
    \subfloat{
      \includegraphics[width=0.48\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
    }
  \end{center}
  \vspace{-0.5cm}
  \caption{
    {\bf Rotation periods for Kepler target stars (left) and known
    planet hosts (right)}.  The periods are from \citet{Santos_2019}
    and \citet{Santos_2021} (\citetalias{Santos_2019} and
    \citetalias{Santos_2021}).  The gray lines are ``mean fits'' from
    Figure~\ref{fig:models} to the slow rotation sequences of M67,
    Ruprecht~147, NGC~6811, Praesepe, and a trio of 120\,Myr open
    clusters.  The stellar sample includes only apparently single stars
    near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
    and 3800$<$$T_{\rm eff}$/K$<$6200.  The displayed planets require
    the same stellar cuts, and include only the confirmed and candidate planets
    described in Section~\ref{subsec:plflags}.
  }
  \label{fig:prot_vs_teff}
\end{figure*}

\begin{figure}[!t]
  \begin{center}
    \leavevmode
    \includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_JUMP_eagles_logy.pdf}
  \end{center}
  \vspace{-0.6cm}
  \caption{
    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
      for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected between 2009 and 2024 comprising
    \nhireshours\,hours of open-shutter time.  Lines are the ``mean''
    isochrones from the same models as in Figure~\ref{fig:models}; the
    intrinsic dispersion beyond $\approx$1\,Gyr becomes much larger
    than changes in the mean.  The logarithmic scale highlights the
    dynamic range of the detections and upper limits.  Some stars with
    lithium detections do not have detected rotation periods, and
    vice-versa.
    \label{fig:li_vs_teff}
  }
\end{figure}

@@

Analyzing the compilation of \citetalias{Santos_2019} and
\citetalias{Santos_2021} rotation periods for the KOI hosts, we
noticed that some known KOIs with rotation periods were missing.
This is not surprising, in that the rotation periods of KOIs have
received considerably more scrutiny than those of ordinary Kepler
stars.
We therefore decided to split our subsequent analysis into a
homogeneous portion that used only the \citetalias{Santos_2019} and
\citetalias{Santos_2021} data, and an inhomogeneous portion that also
considered a broader set of available KOI rotation periods.
For the latter portion, we first included 32 KOIs with orbital and
rotation periods within $\approx$20\% that had been
excluded\footnote{Excluding stars in this category would impose a bias
against young close-in planets, due to the natural commensurability
between the rotation rates of young stars, and Kepler's increased
sensitivity to short-period objects.  Although some of these KOIs are
known false positives, our analysis excludes such objects at later
steps.}%  As one motivating example, this subset of stars includes {\it
%e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
%$P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
%based on membership in RSG-5 \citep{Bouma_2022b}.  }
from the
\citetalias{Santos_2019} and \citetalias{Santos_2021} catalogs
(A.~Santos, private communication).
We then incorporated an additional \nnewdavidtwentyone\ rotation
periods for KOI hosts that
\citet{David_2021} had described as either ``reliable'' or ``highly
reliable'' in their visual analysis of previously reported KOI
rotation periods from \citet{McQuillan_2013,Walkowicz_2013,Mazeh_2015}
and \citet{Angus_2018}.
Inclusion of these additional KOI rotation periods is a
supplementary measure aimed at completeness in our final KOI age
catalog; the provenances of the individual adopted periods are all
noted in the relevant tables.

@@

We considered planets in the NASA Exoplanet Archive's cumulative KOI
table, which included the best knowledge available on any given planet
candidate as of 6 June 2023, while also incorporating human-based
vetting.\footnote{See
\url{https://exoplanetarchive.ipac.caltech.edu/docs/PurposeOfKOITable.html}}
These planets represent a superset of those in the fully automated
Q1-Q17 DR25 KOI Table \citep{Thompson_2018}, which could be adopted in
future work for planet occurrence rate calculations.  This version of
the cumulative KOI table included \nkoisnofp\ objects that are either
``confirmed'' or ``candidate'' planets, after excluding known false
positives.

@@

{\it Kepler-Gaia crossmatch quality (bits 4 and 5)}--- In situations
where we wanted to leverage Gaia DR3 data, we used M.~Bedell's 4$''$
Kepler-to-Gaia crossmatch\footnote{\url{https://gaia-kepler.fun}, last
accessed 15 April 2024.  We used 4$''$ rather than the 1$''$ match
because 91 KOIs are omitted in the 1$''$ match, 23 of which have
either ``candidate'' or ``confirmed'' status.  Almost all of these 23
are M dwarfs with high proper motions.  Performing the same crossmatch
at 4 arcseconds, only two KOIs (KOI-3993 and KOI-6531), both known
false positives, fail to yield Gaia DR3 crossmatches.}, which is a
crossmatch of the NASA Exoplanet Archive
\texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3.  The separation
distribution of the Kepler-Gaia DR3 crossmatches is such that 99.2\%
of candidate matches are within 1$''$.   We nonetheless noted an
upturn in the candidate match rate from 3-4$''$; such sources are
flagged using bit 4.  For KIC stars with multiple potential Gaia
matches within the 4$''$ search radius, we adopted the brightest star
as the default match.  In most such cases this was unambiguous because
there is a large brightness difference between the primary and any
apparent neighbors.  However cases with multiple stars within 4$''$
within $\Delta G$$<$$0.5$\,mag are noted using bit flag 5.

@@

\begin{figure*}[!t]
  \begin{center}
    \includegraphics[width=0.7\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
  \end{center}
  \vspace{-0.5cm}

  \caption{
    {\bf Sizes, orbital periods, and rotation-based ages of transiting
    exoplanets younger than one billion years.} Stars denote
    \nplyounggyrotwosigmanograzingnoruwe\ planets that are newly
    characterized in this work to have $t_{\rm gyro}$<1\,Gyr at
    $2$$\sigma$; circles are planets from the literature meeting the
    same age requirement, but with a heterogeneous set of age
    provenances.  Gray points are transiting planets from the NASA
    Exoplanet Archive older than 1\,Gyr.  If one were to select for
    precise ages ($t/\sigma_{\rm t}>3$), this would hide $\approx$ten
    $<$0.5\,Gyr planets, and add $\approx$fifteen 0.5-1\,Gyr planets.
    For this plot, we require our Kepler stellar hosts to have no
    quality flags raised, and for the planets to
    similarly no have quality flags raised. 
    \label{fig:rp_period_age_results}
  }
\end{figure*}

@@

\begin{enumerate}
  \item Two-sided posteriors available for both $t_{\rm gyro}$ and
    $t_{\rm Li}$: if their values were consistent within 2$\sigma$, we
    judged them to be consistent; if they were consistent within
    2-3$\sigma$, we suggest them to be ``maybe'' consistent.
    Otherwise, they are in disagreement.
    %
  \item If $t_{\rm Li}$ is a lower limit, we compare the $t_{\rm Li}$
    lower limit with the $1\sigma$ upper limit from $t_{\rm gyro}$; if
    the two overlapped, we judge the age estimates to be consistent.
    %
  \item If $t_{\rm Li}$ provides a two-sided posterior, and no
    rotation is reported, we judge the age estimates to be
    inconsistent.  This is because this lithium constraint can only be
    provided for G and K stars $\lesssim$2\,Gyr old, and Kepler should
    have been sensitive to the rotation periods of such stars.
\end{enumerate}

@@

With respect to our original questions, we have 
three main conclusions.
\begin{enumerate}
  %FIXME: 120 number
  \item From the rotation rates of Kepler planet-hosting stars, we
    have identified  \nplyounggyrotwosigma\ planets younger than
    1\,Gyr at 2$\sigma$, and \nplyounggyro\ planets with median ages
    below 1\,Gyr.
    This implies that there are now $\approx$120-170 known
    sub-gigayear planets with well-constrained ages.
  %
  \item The age distributions of both the Kepler
    target stars and the known Kepler planet-hosts shows a
    demographic cliff.  
    There are \ratioobtoybstars\ times as many ``old'' (2-3\,Gyr)
    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.
    The star formation rate today is \ratiosfr$\pm$\uncratiosfr\ times
    lower than it was three billion years ago.
  %
  \item The implication of the latter age distribution is that rather
    than expecting $\approx$500 sub-gigayear exoplanets, we should
    instead expect $\approx$250.
    We have therefore gone from a factor of ten discrepancy to a
    factor of at most two.
%    The issue is therefore mostly, although perhaps not entirely, resolved.
%    Some of the remaining missing young
%    transiting planets may be in the Kepler data.  Many more are
%    likely in K2 and TESS.
\end{enumerate}

>>>>

\author{coauthors, possibly including but not limited to
Hillenbrand, Howard, Isaacson, Palumbo}

% <250 words
\begin{abstract}
  %Recent analyses of FGK stars in open clusters have revealed how
  %stellar rotation rates and lithium content evolve over the first few
  %billion years.
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium
  is measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid; these rotation-based ages are
  roughly an order of magnitude more precise than isochrone ages in
  this sample of FGK stars.
  %
  The independent lithium ages agree with the rotation-based ages for
  $\approx$90\% of cases, suggesting that most of the rotation-based
  ages are not only precise, they are also accurate.
  %However, the timescales of lithium depletion are such that
  %rotation-based ages are more precise for most of our stellar sample.
  %
  Our resulting catalog includes \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  More broadly, we find that the star formation rate in the Kepler
  field has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the
  past 3\,Gyr, equally for both planet hosts and for the overall
  stellar sample.
  %
  This ``demographic cliff'' is connected to the star formation
  history of the Galaxy, and it helps clarify the relative balance of
  young and old planets that have so far been detected using transits.
  %
\end{abstract}

@@

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
billions of years old, similar to the Earth.  This fact leaves many
gaps in our knowledge of what exact dynamical and physical processes
produce these objects.    The reason is that
processes such as thermal cooling \citep{2007ApJ...659.1661F},
atmospheric loss \citep{2019AREPS..47...67O}, giant impacts
\citep{2014prpl.conf..595R}, and dynamical instabilities
\citep{2017MNRAS.470.1750I} are expected to be most efficient over
$\approx$10\,Myr to $\approx$1\,Gyr timescales; when we observe most
known exoplanets, these processes have completed, and not much is happening.

@@

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 0.5-4\,Gyr, and can also provide useful age limits at earlier
times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
clock is quite old
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  
In the context of Kepler,
empirical models have been previously applied to derive rotation-based
ages for subsets of the Kepler data
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.
There have been major improvements since these pioneering studies in:
\begin{enumerate}[label={\it \roman*)},leftmargin=*]
  \item The extent of the Kepler Object of Interest catalog, as well
    as our understanding of which of these objects are false
    positives;
  %
  \item Our empirical understanding of how rotation periods of stars
    in open clusters evolve
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item Our ability to model not only the mean evolution but also the
    intrinsic dispersion in rotation periods as a function of time
    \citep{Bouma_2023}.  
    For Sun-like stars, the new model is broadly consistent with older
    calibrations \citep[e.g.][]{Mamajek_2008}.  However, for late G
    dwarfs and K dwarfs, the age-calibration differs by factors of a
    few;
  %
  \item Our ability to identify binary stars, which can potentially
    problematic for estimates of rotation-based ages.
\end{enumerate}
Today, rotation-based ages for FGK stars below $\approx$4\,Gyr are
empirically tied to the cluster age scale and rotation sequence, and
are limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.

@@

\begin{figure*}[!t]
  \begin{center}
    \subfloat{
      \includegraphics[width=0.49\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
    }
    \subfloat{
    \includegraphics[width=0.49\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
    }
  \end{center}
  \vspace{-0.5cm}
  \caption{
    {\bf The stars.}  Our analysis focuses on stars observed by Kepler
    with previously reported rotation periods (dark gray points).
    The rotation periods were primarily measured by \citet{Santos_2019} and \citet{Santos_2021}.
    About half of these rotators are suitable for gyrochronology
    (black points), based on factors including their temperatures,
    surface gravities, non-binarity, and proximity to the main
    sequence (see Section~\ref{subsec:flags}).   Some are
    ``confirmed'' or ``candidate'' Kepler Objects of Interest that
    meet an additional set of planetary quality criteria (blue points;
    Section~\ref{subsec:plflags}).  Surface gravities and effective
    temperatures were derived photometrically by \citet{Berger_2020a_catalog}.
  }
  \label{fig:stellarprops}
\end{figure*}

\begin{figure*}[!t]
  \begin{center}
    \leavevmode
    \subfloat{
      \includegraphics[width=0.49\textwidth]{gyromodeldispersion.pdf}
    }
    \subfloat{
      \includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf}
    }
  \end{center}
  \vspace{-0.6cm}
  \caption{
    {\bf The models.}
    Points represent $10^4$ draws from model probability distributions
    that have been fitted to rotation periods
    \citep{Bouma_2023} and lithium equivalent widths
    \citep[EWs;][]{Jeffries_2023} of stars in open clusters.  Lines
    are the ``mean models'' at various ages; the asymmetric intrinsic
    dispersion in rotation and lithium about these mean models, which
    is what the models fit, sets the theoretical precision floor for
    both age-dating methods.    Additional sources of uncertainty,
    including measurement uncertainty, impose further limits on
    achievable age precision.  These models are calibrated using open
    clusters younger than 4\,Gyr.  The displayed points assume a
    uniform distribution in temperature for visual clarity.  The sizes
    of the points are the same in each panel, so that low apparent
    density signifies greater dispersion around the mean.
    \label{fig:models}
  }
\end{figure*}

\begin{figure*}[!t]
  \begin{center}
    \subfloat{
      \includegraphics[width=0.48\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
    }
    \subfloat{
      \includegraphics[width=0.48\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
    }
  \end{center}
  \vspace{-0.5cm}
  \caption{
    {\bf Rotation periods for Kepler target stars (left) and known
    planet hosts (right)}.  As in Figure~\ref{fig:stellarprops}, periods are from \citet{Santos_2019}
    and \citet{Santos_2021} (\citetalias{Santos_2019} and
    \citetalias{Santos_2021}).  The gray lines are ``mean fits'' 
    to the slow rotation sequences of M67,
    Ruprecht~147, NGC~6811, Praesepe, and a trio of 120\,Myr open
    clusters.  The stellar sample includes only apparently single stars
    near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
    and 3800$<$$T_{\rm eff}$/K$<$6200.  The displayed planets require
    the same stellar cuts, and include only the confirmed and candidate planets
    described in Section~\ref{subsec:plflags}.
    An order of magnitude estimate for the number of young planets discovered
    by Kepler follows by counting planets below the mean rotational isochrones.
  }
  \label{fig:prot_vs_teff}
\end{figure*}

\begin{figure}[!t]
  \begin{center}
    \leavevmode
    \includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_JUMP_eagles_logy.pdf}
  \end{center}
  \vspace{-0.6cm}
  \caption{
    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
      for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected between 2009 and 2024 that comprised
    \nhireshours\,hours of open-shutter time.  Lines are the ``mean''
    isochrones from the same models as in Figure~\ref{fig:models}; the
    intrinsic dispersion beyond $\approx$1\,Gyr becomes much larger
    than changes in the mean.  The logarithmic scale highlights the
    dynamic range of the detections and upper limits.  Some stars with
    lithium detections do not have detected rotation periods, and
    vice-versa.
    \label{fig:li_vs_teff}
  }
\end{figure}

@@

Analyzing the compilation of \citetalias{Santos_2019} and
\citetalias{Santos_2021} rotation periods for the KOI hosts, we
noticed that some known KOIs with rotation periods were missing.
This is not surprising, in that the rotation periods of KOIs have
received considerably more scrutiny than those of ordinary Kepler
stars.
We therefore decided to split our subsequent analysis into a
homogeneous portion that used only the \citetalias{Santos_2019} and
\citetalias{Santos_2021} data, and an inhomogeneous portion that also
considered a broader set of available KOI rotation periods.
For the latter portion, we first included 32 KOIs with orbital and
rotation periods within $\approx$20\% that had been
excluded %\footnote{Excluding stars in this category would impose a bias
%against young close-in planets, due to the natural commensurability
%between the rotation rates of young stars, and Kepler's increased
%sensitivity to short-period objects.  Although some of these KOIs are
%known false positives, our analysis excludes such objects at later
%steps.}%  As one motivating example, this subset of stars includes {\it
%e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
%$P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
%based on membership in RSG-5 \citep{Bouma_2022b}.  }
from the
\citetalias{Santos_2019} and \citetalias{Santos_2021} catalogs
(A.~Santos, private communication).
We then incorporated an additional \nnewdavidtwentyone\ rotation
periods for KOI hosts that
\citet{David_2021} had described as either ``reliable'' or ``highly
reliable'' in their visual analysis of previously reported KOI
rotation periods from \citet{McQuillan_2013,Walkowicz_2013,Mazeh_2015}
and \citet{Angus_2018}.
Inclusion of these additional KOI rotation periods is a
supplementary measure aimed at completeness in our final KOI age
catalog; the provenances of the individual adopted periods are all
noted in the relevant tables.

@@

We considered planets in the NASA Exoplanet Archive's cumulative KOI
table, which included the best knowledge available on any given planet
candidate as of 6 June 2023, while also incorporating human-based
vetting. %\footnote{See
%\url{https://exoplanetarchive.ipac.caltech.edu/docs/PurposeOfKOITable.html}}
These planets represent a superset of those in the fully automated
Q1-Q17 DR25 KOI Table \citep{Thompson_2018}, which could be adopted in
future work for planet occurrence rate calculations.  This version of
the cumulative KOI table included \nkoisnofp\ objects that are either
``confirmed'' or ``candidate'' planets, after excluding known false
positives.

@@

{\it Kepler-Gaia crossmatch quality (bits 4 and 5)}--- In situations
where we wanted to leverage Gaia DR3 data, we used M.~Bedell's 4$''$
Kepler-to-Gaia crossmatch\footnote{\url{https://gaia-kepler.fun}, last
accessed 15 April 2024.  We used 4$''$ rather than the 1$''$ match
because 91 KOIs are omitted in the 1$''$ match, 23 of which have
either ``candidate'' or ``confirmed'' status.  Almost all of these 23
are M dwarfs with high proper motions.  Performing the same crossmatch
at 4$''$, only two KOIs (KOI-3993 and KOI-6531), both known
false positives, fail to yield Gaia DR3 crossmatches.}, which is a
crossmatch of the NASA Exoplanet Archive
\texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3.  The separation
distribution of the Kepler-Gaia DR3 crossmatches is such that 99.2\%
of candidate matches are within 1$''$.   We nonetheless noted an
upturn in the candidate match rate from 3-4$''$; such sources are
flagged using bit 4.  For KIC stars with multiple potential Gaia
matches within the 4$''$ search radius, we adopted the brightest star
as the default match.  In most such cases this was unambiguous because
there is a large brightness difference between the primary and any
apparent neighbors.  However cases with multiple stars within 4$''$
within $\Delta G$$<$$0.5$\,mag are noted using bit flag 5.

@@

\begin{figure*}[!t]
  \begin{center}
    \includegraphics[width=0.7\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
  \end{center}
  \vspace{-0.5cm}
  \caption{
    {\bf Sizes, orbital periods, and rotation-based ages of transiting
    exoplanets younger than one billion years.} Stars denote
    \nplyounggyrotwosigmanograzingnoruwe\ planets that are newly
    characterized in this work to have $t_{\rm gyro}$<1\,Gyr at
    $2$$\sigma$; circles are planets from the literature meeting the
    same age requirement, but with a heterogeneous set of age
    provenances.  Gray points are transiting planets from the NASA
    Exoplanet Archive older than 1\,Gyr.  If one were to select for
    precise ages ($t/\sigma_{\rm t}>3$), this would hide $\approx$ten
    $<$0.5\,Gyr planets, and add $\approx$fifteen 0.5-1\,Gyr planets.
    For this plot, we require our Kepler stellar hosts to have no
    quality flags raised, and for the planets to
    similarly no have quality flags raised. 
    \label{fig:rp_period_age_results}
  }
\end{figure*}

@@

\begin{enumerate}[leftmargin=*]
  \item Two-sided posteriors available for both $t_{\rm gyro}$ and
    $t_{\rm Li}$: if their values were consistent within 2$\sigma$, we
    judged them to be consistent; if they were consistent within
    2-3$\sigma$, we suggest them to be ``maybe'' consistent.
    Otherwise, they are in disagreement.
    %
  \item If $t_{\rm Li}$ is a lower limit, we compare the $t_{\rm Li}$
    lower limit with the $1\sigma$ upper limit from $t_{\rm gyro}$; if
    the two overlapped, we judge the age estimates to be consistent.
    %
  \item If $t_{\rm Li}$ provides a two-sided posterior, and no
    rotation is reported, we judge the age estimates to be
    inconsistent.  This is because this lithium constraint can only be
    provided for G and K stars $\lesssim$2\,Gyr old, and Kepler should
    have been sensitive to the rotation periods of such stars.
\end{enumerate}

@@

With respect to our original questions, we have 
three main conclusions.
\begin{enumerate}[leftmargin=*]
  %FIXME: 120 number
  \item From the rotation rates of Kepler planet-hosting stars, we
    have identified  \nplyounggyrotwosigma\ planets younger than
    1\,Gyr at 2$\sigma$, and \nplyounggyro\ planets with median ages
    below 1\,Gyr.
    This implies that there are now $\approx$120-170 known
    sub-gigayear planets with well-constrained ages.
  %
  \item The age distributions of both the Kepler
    target stars and the known Kepler planet-hosts shows a
    demographic cliff.  
    There are \ratioobtoybstars\ times as many ``old'' (2-3\,Gyr)
    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.
    The star formation rate today is \ratiosfr$\pm$\uncratiosfr\ times
    lower than it was three billion years ago.
  %
  \item The implication of the latter age distribution is that rather
    than expecting $\approx$500 sub-gigayear exoplanets, we should
    instead expect $\approx$250.
    We have therefore gone from a factor of ten discrepancy to a
    factor of at most two.
%    The issue is therefore mostly, although perhaps not entirely, resolved.
%    Some of the remaining missing young
%    transiting planets may be in the Kepler data.  Many more are
%    likely in K2 and TESS.
\end{enumerate}
